The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel {\normalsize \ldots after that it gets a little tricky.}

### 2013-03-16 trunk/tasbot/paper/paper.tex
Inner loop

performance of the algorithm! I had spent much time tuning the {\tt

@@

\parameteralert{Here I choose the magic number 10 for the size of input
  motifs. On a scale from gravitational constant to pulled it out of
  my ass, this is an 8. We could perhaps justify 10 as being close to
  the speed of input change actually possible for a human (6 button
  presses per second; 166ms). I believe it is possible to do much
  better here and the code contains a few such false starts, but using
  motifs was one of the biggest immediate improvements in the history
  of this code, so I went with it. A natural thing to try is a Markov
  model, although this has a free parameter too (the number of states
  of history). It is likely possible to do some kid of abstract
  interpretation where multiple different input sequences with
  equivalent outcomes are explored simultaneously, which might obviate
  the need for computing an input model from the observed play. The
  playfun algorithm below takes motifs as primitive because of the way
  it was developed; I'll use footnotes to describe my thinking about
  how to remove this.}

@@

In this section I'll describe how the objective functions are used to
play the game. On a scale from canonical to Star Wars Christmas
Special, this algorithm is an 7. So, rather than focus on the
particulars of some of the heuristics, I'll try to give a story of the
different things I tried, what motivated the ideas in the current
version, and my intuitions about what is working well (or not) and
why. This algorithm is called {\tt playfun} and it can be found
implemented in C++ in {\tt playfun.cc}.

@@

In order to use the emulator to search for good sequences of inputs, I
needed deeper integration than just observing memory. The FCEUX
emulator is about a jillion lines of C++-ish code, was intended as an
interactive GUI application, contains support for multiple different
platforms, and the code is, on a scale from a pile of horse shit to
not horse shit, approximately a 2.\footnote{It is, however, an
  excellent emulator to use, has fancy tools for recording and editing
  movies, and is popular in the speedrun community; just don't read
  the code.} With a medium amount of suffering I got the emulator
compiling under mingw in 64-bit mode, and working behind a streamlined
interface ({\tt emulator.h}). Once a game is initialized, it is always
at an input frame---you can give it an 8-bit input (for the
1\st\ player) to step a single frame, or read the 2048 bytes of
memory. You can also save the complete state of the emulator into a
vector of bytes, which allows you to restore the state to exactly that
same point.\footnote{This contains the RAM, but also stuff we didn't
  consider, like the Picture Processing Unit's state, and internal
  emulator stuff.} These save-states are portable across sessions as
long as the code was compiled and the game initialized the right
way.\footnote{The original FCEUX supports portable save-states, but I
  removed that guarantee in order to get the smallest possible byte
  vectors. More on that below.} FCEUX must be single-threaded because
it uses global variables galore. I made several enhancements to the
emulator interface, which are discussed later.

It's important to know that almost all the CPU time in all the
algorithms discussed in this paper is spent emulating NES frames; it
takes about 500\textmu s to process a single step. Lots of
engineering effort goes into reducing the number of frames the
algorithms emulate.

\subsection{Naive attempts}

@@

I wish this worked, because that would be truly laughable, but
on a scale from doesn't to does it's a 1. At best, Mario twitches
in place. The inputs that it plays are insane. There are lots of
reasons, but a clear one is that a single input rarely affects
your progress in the game on the very next frame. I didn't really
expect this approach to work and I already knew that the state
space is too big to search exhaustively, which is why I implemented
motifs. This drastically reduces the search space and makes each
step more significant; the inner loop can now be:

@@

The flipside of avoiding danger is seeking adventure. Mario can avoid
danger for quite a long time by just waiting at the beginning of the
game, until time runs out. He can dilly dally before a pit,
contemplating the void. But princesses need rescuin'. The same
approach as before works for evaluating a state in terms of its
potential: Try some random futures and look at where we end up. We
could take the max over those scores if we're being optimistic, or the
average or sum if we're trying to judge the general value of the
state. In my first version, which did work pretty well, I took the
max; so basically I had the min of some random states plus the max of
some other states. But why generate a bunch of futures and take the
min of some and the max of some others? On a scale of Thank You Mario
Your Quest Is Over We Present You A New Quest Push Button B to I'm
Sorry, but the Princess is in Another Similarly-Shaped but Not Exactly
that Samesuch Castle, this is an 8.

@@

Taking the minimum of random futures is always silly (at least if we
believe our objective function), because nothing other than our own
bad memory can force us to take a series of steps if we know a
different better series of steps. Taking the max is possibly foolish
if we don't know how to reconstruct a rare good state that caused the
maximum score to be high. Both lead us to the idea: Keep around
a set of candidate futures that we use for evaluation and search,
rather than just randomly generating futures when we want to evaluate
a state. This turns out to work really well and be more efficient.

HERE

@@

% good parts!
% pacman @6800

>>>>

quality of the output! I had spent much time tuning the {\tt

@@

\parameteralert{Here I choose the magic number 10 for the size of input
  motifs. On a scale from gravitational constant to pulled it out of
  my ass, this is an 8. We could perhaps justify 10 as being close to
  the speed of input change actually possible for a human (6 button
  presses per second; 166ms). I believe it is possible to do much
  better here and the code contains a few such false starts, but using
  motifs was one of the biggest immediate improvements in the history
  of this code, so I went with it. A natural thing to try is a Markov
  model, although this has a free parameter too (the number of states
  of history). It is likely possible to do some kind of abstract
  interpretation where multiple different input sequences with
  equivalent outcomes are explored simultaneously, which might obviate
  the need for computing an input model from the observed play. The
  playfun algorithm below takes motifs as primitive because of the way
  it was developed; I'll use footnotes to describe my thinking about
  how to remove this.}

@@

In this section I'll describe how the objective functions are used to
play the game. On a scale from canonical to Star Wars Christmas
Special, this algorithm is an 7. So, rather than focus on the
particulars of some of the heuristics, I'll try to give a story of the
different things I tried, what motivated the ideas in the current
version, and my intuitions about what is working well (or not) and
why. This algorithm is called {\tt playfun} and it can be found
implemented in C++ in {\tt playfun.cc}; some historic versions are
in {\tt playfun-*.cc}.

@@

In order to use the emulator to search for good sequences of inputs, I
needed deeper integration than just observing memory. The FCEUX
emulator is about a jillion lines of C++-ish code, was intended as an
interactive GUI application, contains support for multiple different
platforms, and the code is, on a scale from a pile of horse shit to
not horse shit, approximately a 2.\footnote{It is, however, an
  excellent emulator to use, has fancy tools for recording and editing
  movies, and is popular in the speedrun community. I highly recommend
  it; just don't read the code.} With a medium amount of suffering I
got the emulator compiling under mingw in 64-bit mode, and working
behind a streamlined interface ({\tt emulator.h}). Once a game is
initialized, it is always at an input frame---you can give it an 8-bit
input (for the 1\st\ player) to step a single frame, or read the 2048
bytes of memory. You can also save the complete state of the emulator
into a vector of bytes, which allows you to restore the state to
exactly that same point.\footnote{This contains the RAM, but also
  stuff we didn't consider, like registers, the Picture Processing
  Unit's state, and internal emulator stuff.} These save-states are
portable across sessions as long as the code was compiled and the game
initialized the right way.\footnote{The original FCEUX supports
  portable save-states, but I removed that guarantee in order to get
  the smallest possible byte vectors. More on that below.} FCEUX must
be single-threaded because it uses global variables galore. I made
several enhancements to the emulator interface, which are discussed
later.

It's important to know that almost all the CPU time in all the
algorithms discussed in this paper is spent emulating NES frames; it
takes about 500\textmu s to process a single step. Lots of engineering
effort goes into reducing the number of frames the algorithms emulate.
The {\tt playfun} program takes a ROM file, the learned objectives and
motifs, and runs on the console for arbitrarily long, outputting a
movie file consisting of the input sequence it think is best. The
current {\tt playfun} algorithm is much too slow to run real-time, but
it would be possible to have video output as it played. I disabled
most of the video code, however, in an effort to make the emulation
loop run as fast as possible.

\subsection{Naive attempts} \label{sec:naive}

@@

I wish this worked, because that would be truly laughable (and is fast
enough to run real-time), but on a scale from doesn't to does it's a
1. At best, Mario twitches in place. The inputs that it plays are
insane. There are lots of reasons, but a clear one is that a single
input rarely affects your progress in the game on the very next frame.
I didn't really expect this approach to work and I already knew that
the state space is too big to search exhaustively, which is why I
implemented motifs. This drastically reduces the search space and
makes each step more significant; the inner loop can now be:

@@

The flipside of avoiding danger is seeking adventure. Mario can avoid
danger for quite a long time by just waiting at the beginning of the
game, until time runs out. He can dilly dally before a pit,
contemplating the void. But princesses need rescuin'. The same
approach as before works for evaluating a state in terms of its
potential: Try some random futures and look at where we end up. We
could take the max over those scores if we're being optimistic, or the
average or sum if we're trying to judge the general value of the
state. In my first version, which did work pretty well, I took the
max; so basically I had the min of some random states plus the max of
some other states. But why generate a bunch of futures and take the
min of some and the max of some others? On a scale of Thank You Mario
Your Quest Is Over We Present You A New Quest Push Button B to I'm
Sorry, but Our Princess is in Another Similarly-Shaped but Not Exactly
that Samesuch Castle, this is an 8.

@@

Taking the minimum of random futures is always silly (at least if we
believe our objective function), because nothing other than our own
bad memory can force us to take a series of steps if we know a
different better series of steps. Taking the max is possibly foolish
if we don't know how to reconstruct a rare good state that caused the
maximum score to be high. Both lead us to the idea: Keep around a set
of candidate futures that we use for evaluation and search, rather
than just randomly generating futures when we want to evaluate a state
and then discarding them. This turns out to work really well and be
more efficient.

The basic version of the {\tt playfun} algorithm looks like this.
Maintain {\tt NFUTURES} futures (this is 40 for the results presented
in Section~\ref{sec:results}), originally just seeded with weighted
random motifs. We aren't likely to execute any of these futures
verbatim, but they are intended to give us high watermarks of what
we're capable of, plus allow us to judge future danger. As we execute
search, futures will be partly consumed and extended, and some discarded
and replaced.

Each future stores a desired length from 50--800 frames, and whenever
it is shorter than that, we extend it (at its end) with random weighted
motifs. The inner pseudoloop then looks like this:

\begin{code}
for (;;) \{
  vector<uint8> before = GetMemory();
  vector<uint8> state = GetState();
\
  set<vector<uint8>> nexts;
  for (Future f : futures) \{
   nexts.insert(f.First10Frames());
   f.ChopOffFirst10Frames();
  \}
\ 
  while (nexts.size() < NFUTURES)
   nexts.push\_back(/* random motif */);
\ 
  for (vector<uint8> &next : nexts) \{
    RestoreState(state);
    for (uint8 i : next) Step(i);
    double score = ScoreByFutures(before, futures);
    // Save the best-scoring next...
    // 
  \}
\
  ReplaceBadFutures(futures);
  ExtendFuturesToDesiredLengths(futures);
\}
\end{code}

At each iteration of the loop we will find the best next 10-frame
sequence to commit to. Rather than search over all motifs, we search
over the first 10 frames of each future. This has several important
consequences. First, since futures are generated by weighted motifs,
it makes it more likely that we spend CPU time evaluating motifs that
are common; the code from Section~\ref{sec:naive} always explores
every motif, even rare ones. Second, it guarantees that if we pick
some 10 frames on the basis of a single good future, we don't have to
worry about recreating that future; it will still be among our futures
for the next round. This is key: It allows us to use the determinism
of the game to replay a really good future if we find one, not just
use average random futures to assess how good a state will be.

The function {\tt ScoreByFutures} saves the state, then runs each
of the {\tt NFUTURES} futures to get a final state and memory. We
score each final memory relative to the start memory. Scoring is
a matter of

@@

\subsection{MARIONET}

\subsection{Backtracking}
